Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Guideline for International \emph{and Domestic} Students

### 2024-02-27 demystify.tex
updates

So what can you do about this? You can explain about your school in your statement (ask your LOR writer to do that too). Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS PhD programs in the US that have Vietnamese professors}. It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school that you didn't know much about can have very strong research in your interested topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research finding.}. \autoref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

  I have read your TSE'21 paper on numerical invariant generation, and I am interested in this line of dynamic invariant research. I have worked (optional: with prof. Y at Z) on static program analysis and I think it could be used to tackle the spurious issues mentioned in your paper. I have a small paper at conference/workshop C and a project on symbolic execution at Github G.

@@

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. In fact, if a faculty is not in CS, it is unlikely that they can be the \emph{main} adviser of a CS PhD student. They may \emph{co-advise} or be in the PhD dissertation committee, but your main adviser will need to be a tenured or tenure-track faculty in CS. If in doubt, check with the CS department for their requirements.  For this specific reason,  CSRankings includes only tenure line faculty who can advise CS PhD students. I also have compiled a \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{list} of Vietnamese faculty who can advise PhD students. \autoref{sec:faculty-types} talks more about who can serve as your PhD adviser.

@@

\section{Differences between PhD in the US and Other Countries}\label{sec:non-us-differences}

@@

Students, especially international ones, often get confused on what to do after they get an offer from a professor, e.g., how to postpone or accept/reject offers, and most importantly, how to do so without offending anyone. In short, if you are not serious about an offer for any reason (e.g., still waiting for other offers), do not accept, i.e., commit to it. If you are not sure, ask for more time to decide.  But do not accept an offer and then later reject it.  This is considered \emph{rude and unprofessional}, and will likely burn bridges with that professor.

\paragraph{Accepting an offer} If you decide to accept an offer, you can say: \emph{``Thank you for the offer.  I would like to accept it and look forward to working with you.  Could you please send me more details about the offer and what to do next?''}.  Here you are not only accepting the offer but also asking for more details. The prof. will likely send you more details about the offer and what to do next.  If you decide to accept an offer, do so quickly.

If you have better offers, you might mention them to the professor and ask if they could match them. It is unlikely they can do so as RA might be fixed by the department and not negotiable. They might recommend you for departmental fellowships or some additional perks.

@@

I started writing this document in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). I have put the source code of this document on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that anyone can contribute to it.

>>>>

What can you do about this? You can explain about your school in your statement (ask your LOR writer to do that too). Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS PhD programs in the US that have Vietnamese professors}. %It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}. You will be very surprised to learn that a school that you didn't know much about can have very strong research in your interested topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research finding.}. \autoref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

  I have read your TSE'21 paper on numerical invariant generation, and I am interested in this line of dynamic invariant research. I have worked (optional: with prof. Y at Z) on static program analysis and I think it could be used to tackle the spurious issues mentioned in your paper. I have a short paper at conference/workshop C and a project on symbolic execution (Github repo G).

@@

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. In fact, if a faculty is not in CS, it is unlikely that they can be the \emph{main} adviser of a CS PhD student. They may \emph{co-advise} or be in the PhD dissertation committee, but your main adviser will likely be a tenured or tenure-track faculty in CS. If in doubt, check with the CS department for their requirements.

For this specific reason,  CSRankings includes only tenure line faculty who can advise CS PhD students. I also have compiled a \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{list} of Vietnamese faculty who can advise PhD students. \autoref{sec:faculty-types} talks more about who can serve as your PhD adviser.

@@

\section{PhDs in the US vs. Other Countries}\label{sec:non-us-differences}

@@

Students, especially international ones, often get confused on what to do after they get an offer from a professor, e.g., how to postpone or accept/reject offers, and most importantly, how to do so without offending anyone. In short, if you are not serious about an offer for any reason (e.g., still waiting for other offers), do not accept it.
If you are not sure, ask for more time to decide.  But do not accept an offer and then later reject it.  This is considered \emph{rude and unprofessional}, and will likely burn bridges with that prof.

\paragraph{Accepting an offer} If you decide to accept an offer, you can say: \emph{``Thank you for the offer.  I would like to accept it and look forward to working with you.  Could you please send me more details about the offer and what to do next?''}. The prof. will likely send you more details about the offer and what to do next.  If you decide to accept an offer, do so quickly.

% If you have better offers, you might mention them to the professor and ask if they could match them. It is unlikely they can do so as RA might be fixed by the department and not negotiable. They might recommend you for departmental fellowships or some additional perks.

@@

I started writing this document in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). I have put the source code of this document on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that others can contribute to it.
